Wallace Session 2021-07-13
Erica Johnson
Cloud Forest VCF thresholded: Carso Huasteco
================

Please find below the R code history from your *Wallace* v1.9.0 session.
### Package installation

Wallace uses the following R packages that must be installed and loaded
before starting.

```{r}
library(spocc)
library(spThin)
library(dismo)
library(rgeos)
library(raster)
library(ENMeval)
library(wallace)
```

## Analysis for *Cloud forest* (Cf)

User CSV path with occurrence data. If the CSV file is not in the
current workspace, change to the correct file path
(e.g. “/Users/darwin/Documents/occs/”).

```{r}
# NOTE: provide the folder path of the .csv file
occs_path <- "data/OccData/VCF_thresholded/CarsoHuasteco"
occs_path <- file.path(occs_path, "CF_Occ_VCF100.csv")
# get a list of species occurrence data
userOccs_Cf <- occs_userOccs(
  txtPath = occs_path, 
  txtName = "CF_Occ_VCF100.csv", 
  txtSep = ",", 
  txtDec = ".")
occs_Cf <- userOccs_Cf$Cloud_forest$cleaned
```

### Obtain environmental data

Using user-specified variables.

```{r}
## Specify the directory with the environmental variables
dir_envs_Cf <- "data/Env/Env_CarsoHuasteco"
envs_path <- file.path(dir_envs_Cf, dir(dir_envs_Cf, pattern = ".tif$"))
# Create environmental object (**)
envs_Cf <- envs_userEnvs(
  rasPath = envs_path,
  rasName = dir(dir_envs_Cf, pattern = ".tif$"),
  doBrick = FALSE)
occs_xy_Cf <- occs_Cf[c('longitude', 'latitude')]
occs_vals_Cf <- as.data.frame(raster::extract(envs_Cf, occs_xy_Cf))
# remove occurrence records with NA environmental values
occs_Cf <- occs_Cf[!(rowSums(is.na(occs_vals_Cf)) > 1), ]
# also remove variable value rows with NA environmental values
occs_vals_Cf <- na.omit(occs_vals_Cf)
# add columns for env variable values for each occurrence record
occs_Cf <- cbind(occs_Cf, occs_vals_Cf)
```

### Process environmental data

Sampling the entire available environmental background using a “point buffers” method with a 0.5 degree buffer.

```{r}
# Generate background extent 
bgExt_Cf <- penvs_bgExtent(
  occs = occs_Cf,
  bgSel = "point buffers",
  bgBuf = 0.5)
# Mask environmental data to provided extent
bgMask_Cf <- penvs_bgMask(
  occs = occs_Cf,
  envs = envs_Cf,
  bgExt = bgExt_Cf)
# Sample the full environmental background from the provided area
bgSample_Cf <- penvs_bgSample(
  occs = occs_Cf,
  bgMask =  bgMask_Cf,
  bgPtsNum = sum(!is.na(getValues(bgMask_Cf[[1]]))))
# Extract values of environmental layers for each background point
bgEnvsVals_Cf <- as.data.frame(raster::extract(bgMask_Cf,  bgSample_Cf))
##Add extracted values to background points table
bgEnvsVals_Cf <- cbind(scientific_name = paste0("bg_", "Cloud forest"), bgSample_Cf,
                            occID = NA, year = NA, institution_code = NA, country = NA,
                            state_province = NA, locality = NA, elevation = NA,
                            record_type = NA, bgEnvsVals_Cf)
```

### Partition occurrence data

Partition occurrences and background points for model training and
validation using “spatial block”, a spatial partition method with an
aggregation factor of 2.

```{r}
# R code to get partitioned data
groups_Cf <- part_partitionOccs(
  occs = occs_Cf ,
  bg =  bgSample_Cf, 
  method = "block",
  bgMask = bgMask_Cf,
  aggFact = 2) 
```

### Build and Evaluate Niche Model

Generating a species distribution model using the maxent.jar algorithm
as implemented in ENMeval V2.0 (with clamping = FALSE). For tuning using
L, LQ, H, LQH feature classes and regularization multipliers in the 1, 5
range increasing by 1. Using a total of 1 categorical predictor
variables.

```{r}
# Run maxent model for the selected species
# , 'LQ','H','LQH'
model_Cf <- model_maxent(
  occs = occs_Cf,
  bg = bgEnvsVals_Cf,
  user.grp = groups_Cf, 
  bgMsk = bgMask_Cf,
  rms = c(2.5, 2.5), 
  rmsStep = 0.5,
  fcs = c('L'),
  clampSel = TRUE,
  algMaxent = "maxnet",
  catEnvs = "Soil",
  parallel = TRUE,
  numCores = 4)
```

```{r}
#exporting ENMeval results
dir_eval <- "Outputs/EvalTbl/CarsoHuasteco"
write.csv(model_Cf@results, paste(dir_eval, "CF_VCF100_CH.csv"))
```

### Visualize

Generate a map of the Maxent generated model with no threshold (cloglog).

```{r}
# Select current model and obtain raster prediction
m_Cf <- model_Cf@models[["fc.L_rm.2.5"]]
predSel_Cf <- predictMaxnet(m_Cf, bgMask_Cf, type = "cloglog", clamp = TRUE)
```

## Calculating model thresholds

Calculating the minimum training presence and 10 pct threshold. 

```{r}
pred_vals_Cf <- raster::extract(predSel_Cf, occs_xy_Cf)

MTP <- min(na.omit(pred_vals_Cf))
pct10 <-  rev(sort(pred_vals_Cf))[(ceiling(length(pred_vals_Cf) * 0.9))]

MTP
pct10
```

## Project model

Projecting the model to a new user provided area

```{r}
# Loading the Carso Huasteco shapefile and defining it as the projection area
proj_path <- "data"
proj_userExt_Cf <- rgdal::readOGR(file.path(proj_path, "CarsoHuasteco.shp"), 
                                  layer = "CarsoHuasteco")
  

# Create object of projection variables
projAreaEnvs_Cf <- envs_Cf
# Generate a projection of the model to the desired area
proj_area_Cf <- proj_area(
  evalOut = model_Cf,
  curModel = "fc.L_rm.2.5",
  envs = projAreaEnvs_Cf , 
  outputType = "cloglog",
  alg = "maxnet",
  clamp = TRUE,
  pjExt = proj_userExt_Cf) 
```
Exporting the model projection raster
```{r}
out_path <- "Outputs/RasPreds/CarsoHuasteco"
writeRaster(proj_area_Cf$projArea, 
            paste(out_path, "CF_VCF100.tif", sep = "/"), format = "GTiff")
```

## Model Transfer 

Transferring the model to the Gran Sierra Plegada

```{r}
# Loading the Gran Sierra Plegada shapefile and defining it as the projection area
transf_path <- "data"
transf_userExt <- rgdal::readOGR(file.path(proj_path, "SierraPlegada.shp"), layer = "SierraPlegada")
  

# Create object of projection variables
dir_envs_transf <- "data/Env/Env_SierraPlegada"
envs_transf_path <- file.path(dir_envs_transf, dir(dir_envs_transf, 
                                                   pattern = ".tif$"))
# Create environmental object (**)
envs_transf <- envs_userEnvs(
  rasPath = envs_transf_path,
  rasName = dir(dir_envs_transf, pattern = ".tif$"),
  doBrick = FALSE)
```

```{r}
# Generate a projection of the model to the desired area
transf_area_Cf <- proj_area(
  evalOut = model_Cf,
  curModel = "fc.L_rm.2.5",
  envs = envs_transf, 
  outputType = "cloglog",
  alg = "maxnet",
  clamp = TRUE,
  pjExt = transf_userExt) 
```

Exporting the transferred model raster
```{r}
out_path <- "Outputs/Transfer"
writeRaster(transf_area_Cf$projArea, 
            paste(out_path, "CF_VCF100_CH2GSP.tif", sep = "/"), format = "GTiff")
```
