Wallace Session 2021-11-09
Erica Johnson
Pine-oak Forest Random Filter (25%): Carso Huasteco
================

Please find below the R code history from your *Wallace* v1.9.0 session.

### Package installation

Wallace uses the following R packages that must be installed and loaded
before starting.

```{r}
library(spocc)
library(spThin)
library(dismo)
library(rgeos)
library(raster)
library(ENMeval)
library(wallace)
```

------------------------------------------------------------------------

## Analysis for *Pine-Oak forest* (Pf)

User CSV path with occurrence data. If the CSV file is not in the
current workspace, change to the correct file path
(e.g. “/Users/darwin/Documents/occs/”).

```{r}
# NOTE: provide the folder path of the .csv file
occs_path <- "data/OccData/VCF_RandFilter/CarsoHuasteco"
occs_path <- file.path(occs_path, "PineOak_forest_VCF25.csv")
# get a list of species occurrence data
userOccs_Pf <- occs_userOccs(
  txtPath = occs_path, 
  txtName = "PineOak_forest_VCF25", 
  txtSep = ",", 
  txtDec = ".")
occs_Pf <- userOccs_Pf$PineOak_forest$cleaned
```

### Obtain environmental data

Using user-specified variables.

```{r}
## Specify the directory with the environmental variables
dir_envs_Pf <- "data/Env/Env_CarsoHuasteco"
envs_path <- file.path(dir_envs_Pf, dir(dir_envs_Pf, pattern = ".tif$"))
# Create environmental object (**)
envs_Pf <- envs_userEnvs(
  rasPath = envs_path,
  rasName = dir(dir_envs_Pf, pattern = ".tif$"),
  doBrick = FALSE)
occs_xy_Pf <- occs_Pf[c('longitude', 'latitude')]
occs_vals_Pf <- as.data.frame(raster::extract(envs_Pf, occs_xy_Pf))
# remove occurrence records with NA environmental values
occs_Pf <- occs_Pf[!(rowSums(is.na(occs_vals_Pf)) > 1), ]
# also remove variable value rows with NA environmental values
occs_vals_Pf <- na.omit(occs_vals_Pf)
# add columns for env variable values for each occurrence record
occs_Pf <- cbind(occs_Pf, occs_vals_Pf)
```

### Process environmental data

Sampling the entire available environmental background using a “point buffers” method with a 0.25 degree buffer.

```{r}
# Generate background extent 
bgExt_Pf <- penvs_bgExtent(
  occs = occs_Pf,
  bgSel = "point buffers",
  bgBuf = 0.25)
# Mask environmental data to provided extent
bgMask_Pf <- penvs_bgMask(
  occs = occs_Pf,
  envs = envs_Pf,
  bgExt = bgExt_Pf)
# Sample the full environmental background from the provided area
bgSample_Pf <- penvs_bgSample(
  occs = occs_Pf,
  bgMask =  bgMask_Pf,
  bgPtsNum = sum(!is.na(getValues(bgMask_Pf[[1]]))))#ADDED BY EEJ
# Extract values of environmental layers for each background point
bgEnvsVals_Pf <- as.data.frame(raster::extract(bgMask_Pf,  bgSample_Pf))
##Add extracted values to background points table
bgEnvsVals_Pf <- cbind(scientific_name = paste0("bg_", "PineOak forest"), bgSample_Pf,
                            occID = NA, year = NA, institution_code = NA, country = NA,
                            state_province = NA, locality = NA, elevation = NA,
                            record_type = NA, bgEnvsVals_Pf)
```

### Partition occurrence data

Partition occurrences and background points for model training and
validation using “spatial block”, a spatial partition method with an
aggregation factor of 2.

```{r}
# R code to get partitioned data
groups_Pf <- part_partitionOccs(
  occs = occs_Pf ,
  bg =  bgSample_Pf, 
  method = "block",
  bgMask = bgMask_Pf,
  aggFact = 2) 
```

### Build and Evaluate Niche Model

Generating a species distribution model using the maxent.jar algorithm
as implemented in ENMeval V2.0 (with clamping = FALSE). For tuning using
L, LQ, H, LQH, P, T feature classes and regularization multipliers in the 1, 5
range increasing by 1. Using a total of 1 categorical predictor
variables.

```{r}
# Run maxent model for the selected species
model_Pf <- model_maxent(
  occs = occs_Pf,
  bg = bgEnvsVals_Pf,
  user.grp = groups_Pf, 
  bgMsk = bgMask_Pf,
  rms = c(1,5), 
  rmsStep = 0.5,
  fcs = c('L','LQ','H','LQH','P','T'),
  clampSel = TRUE,
  algMaxent = "maxnet",
  catEnvs = "Soil",
  parallel = TRUE,
  numCores = 4)
```

```{r}
#exporting ENMeval results
dir_eval <- "Outputs/EvalTbl/CarsoHuasteco"
write.csv(model_Pf@results, paste(dir_eval, "PO_VCF25_CH.csv", sep = "/"))
```

### Visualize

Generate a map of the Maxent generated model with no threshold (cloglog).

```{r}
# Select current model and obtain raster prediction
m_Pf <- model_Pf@models[["fc.LQH_rm.1"]]
predSel_Pf <- predictMaxnet(m_Pf, bgMask_Pf, type = "cloglog", clamp = TRUE)
```

Obtaining and exporting lambdas files
```{r}
lambdas_Pf <- lambdasDF(m_Pf, alg = "maxnet")
write.csv(lambdas_Pf, "Outputs/ResponseCurves/CarsoHuasteco/PO_VCF25_lambdas.csv")
```

Creating variable response curves and exporting as a pdf file
```{r, warning = F, echo = F}
pdf(file = "Outputs/ResponseCurves/CarsoHuasteco/PO_VCF25_respcurve.pdf")
lapply(names(bgMask_Pf), maxnet::response.plot, mod = m_Pf, vars = bgMask_Pf, type = "cloglog")

while(!is.null(dev.list())) dev.off()
```
## Calculating model thresholds

Calculating the minimum training presence and 10 pct threshold. 

```{r}
#ADDED TO WALLACE RMD BY EEJ
pred_vals_Pf <- raster::extract(predSel_Pf, occs_xy_Pf)

MTP <- min(na.omit(pred_vals_Pf))
pct10 <-  rev(sort(pred_vals_Pf))[(ceiling(length(pred_vals_Pf) * 0.9))]

MTP
pct10
```

## Project model

Projecting the model to a new user provided area

```{r}
# Loading the Carso Huasteco shapefile and defining it as the projection area
proj_path <- "data"
proj_userExt_Pf <- rgdal::readOGR(file.path(proj_path, "CarsoHuasteco.shp"), 
                                  layer = "CarsoHuasteco")
  

# Create object of projection variables
projAreaEnvs_Pf <- envs_Pf
# Generate a projection of the model to the desired area
proj_area_Pf <- proj_area(
  evalOut = model_Pf,
  curModel = "fc.LQH_rm.1",
  envs = projAreaEnvs_Pf , 
  outputType = "cloglog",
  alg = "maxnet",
  clamp = TRUE,
  pjExt = proj_userExt_Pf) 
```

Exporting the model projection raster
```{r}
out_path <- "Outputs/RasPreds/CarsoHuasteco"
writeRaster(proj_area_Pf$projArea, 
            paste(out_path, "PO_VCF25.tif", sep = "/"), format = "GTiff")
```
