Wallace Session 2021-07-13
Erica Johnson
Submontane Scrubland VCF thresholded: Carso Huasteco
================

Please find below the R code history from your *Wallace* v1.9.0 session.

### Package installation

Wallace uses the following R packages that must be installed and loaded
before starting.

```{r}
library(spocc)
library(spThin)
library(dismo)
library(rgeos)
library(raster)
library(ENMeval)
library(wallace)
```

------------------------------------------------------------------------

## Analysis for *Submontane scrubland* (Ss)

```{r}
# NOTE: provide the folder path of the .csv file
occs_path <- "data/OccData/VCF_thresholded/CarsoHuasteco"
occs_path <- file.path(occs_path, "SS_Occ_VCF100.csv")
# get a list of species occurrence data
userOccs_Ss <- occs_userOccs(
  txtPath = occs_path, 
  txtName = "SS_Occ_VCF100", 
  txtSep = ",", 
  txtDec = ".")
occs_Ss <- userOccs_Ss$Submontane_scrub$cleaned
```

### Obtain environmental data

Using user-specified variables.

```{r}
## Specify the directory with the environmental variables
dir_envs_Ss <- "data/Env/Env_CarsoHuasteco"
envs_path <- file.path(dir_envs_Ss, dir(dir_envs_Ss, pattern = ".tif$"))
# Create environmental object (**)
envs_Ss <- envs_userEnvs(
  rasPath = envs_path,
  rasName = dir(dir_envs_Ss, pattern = ".tif$"),
  doBrick = FALSE)
occs_xy_Ss <- occs_Ss[c('longitude', 'latitude')]
occs_vals_Ss <- as.data.frame(raster::extract(envs_Ss, occs_xy_Ss))
# remove occurrence records with NA environmental values
occs_Ss <- occs_Ss[!(rowSums(is.na(occs_vals_Ss)) > 1), ]
# also remove variable value rows with NA environmental values
occs_vals_Ss <- na.omit(occs_vals_Ss)
# add columns for env variable values for each occurrence record
occs_Ss <- cbind(occs_Ss, occs_vals_Ss)
```

### Process environmental data

Sampling the entire available environmental background using a “point buffers” method with a 0.5 degree buffer.

```{r}
# Generate background extent 
bgExt_Ss <- penvs_bgExtent(
  occs = occs_Ss,
  bgSel = "point buffers",
  bgBuf = 0.5)
# Mask environmental data to provided extent
bgMask_Ss <- penvs_bgMask(
  occs = occs_Ss,
  envs = envs_Ss,
  bgExt = bgExt_Ss)
# Sample the full environmental background from the provided area
bgSample_Ss <- penvs_bgSample(
  occs = occs_Ss,
  bgMask =  bgMask_Ss,
  bgPtsNum = sum(!is.na(getValues(bgMask_Ss[[1]]))))
# Extract values of environmental layers for each background point
bgEnvsVals_Ss <- as.data.frame(raster::extract(bgMask_Ss,  bgSample_Ss))
##Add extracted values to background points table
bgEnvsVals_Ss <- cbind(scientific_name = paste0("bg_", "Submontane scrub"), bgSample_Ss,
                            occID = NA, year = NA, institution_code = NA, country = NA,
                            state_province = NA, locality = NA, elevation = NA,
                            record_type = NA, bgEnvsVals_Ss)
```

### Partition occurrence data

Partition occurrences and background points for model training and
validation using “spatial block”, a spatial partition method with an
aggregation factor of 2.

```{r}
# R code to get partitioned data
groups_Ss <- part_partitionOccs(
  occs = occs_Ss ,
  bg =  bgSample_Ss, 
  method = "block",
  bgMask = bgMask_Ss,
  aggFact = 2) 
```

### Build and Evaluate Niche Model

Generating a species distribution model using the maxent.jar algorithm
as implemented in ENMeval V2.0 (with clamping = FALSE). For tuning using
L, LQ, H, LQH feature classes and regularization multipliers in the 1, 5
range increasing by 1. Using a total of 1 categorical predictor
variables.

```{r}
# Run maxent model for the selected species
model_Ss <- model_maxent(
  occs = occs_Ss,
  bg = bgEnvsVals_Ss,
  user.grp = groups_Ss, 
  bgMsk = bgMask_Ss,
  rms = c(1, 5), 
  rmsStep = 0.5,
  fcs = c('L', 'LQ','H','LQH'),
  clampSel = TRUE,
  algMaxent = "maxnet",
  catEnvs = "Soil_CarsoHuasteco",
  parallel = TRUE,
  numCores = 4)
```

```{r}
#exporting ENMeval results
dir_eval <- "Outputs/EvalTbl/CarsoHuasteco"
write.csv(model_Ss@results, paste(dir_eval, "SS_VCF100_CH.csv", sep = "/"))
```

### Visualize
Generate a map of the Maxent generated model with no threshold (cloglog).

```{r}
# Select current model and obtain raster prediction
m_Ss <- model_Ss@models[["fc.LQH_rm.1.5"]]
predSel_Ss <- predictMaxnet(m_Ss, bgMask_Ss, type = "cloglog", clamp = TRUE)
```

## Calculating model thresholds

Calculating the minimum training presence threshold. 

```{r}
pred_vals_Ss <- raster::extract(predSel_Ss, occs_xy_Ss)
MTP <- min(na.omit(pred_vals_Ss))
MTP
```

## Project model

Projecting the model to a new user provided area

```{r}
# Loading the Carso Huasteco shapefile and defining it as the projection area
proj_path <- "data"
proj_userExt_Ss <- rgdal::readOGR(file.path(proj_path, "CarsoHuasteco.shp"), 
                                  layer = "CarsoHuasteco")
  

# Create object of projection variables
projAreaEnvs_Ss <- envs_Ss
# Generate a projection of the model to the desired area
proj_area_Ss <- proj_area(
  evalOut = model_Ss,
  curModel = "fc.LQ_rm.1.5",
  envs = projAreaEnvs_Ss , 
  outputType = "cloglog",
  alg = "maxnet",
  clamp = TRUE,
  pjExt = proj_userExt_Ss) 
```

Exporting the model projection raster
```{r}
out_path <- "Outputs/RasPreds/CarsoHuasteco"
writeRaster(proj_area_Ss$projArea, 
            paste(out_path, "SS_VCF100.tif", sep = "/"), format = "GTiff")
```